---
title: "Ageing + X bulk mRNAseq (QC)"
format: html
author: Tamas Lakat
---

# Experiment

The dataset contains bulk mRNA-sequencing data obtained from mouse heart samples (Young Control (YC; n=4), Aged Control (AC; n=4), Aged Treated (X; n=10))

# Preprocessing pipeline

1.  **Trimming + QC**
    - Software: fastp
    - Folders: 0_raw ->  1_trimmed
2.  **QC**
    - Software: fastqc, multiqc
    - Folders: 1_trimmed -> 2_fastqc
3.  **Pseudoalignment**
    - Software: Kallisto
    - Reference transcriptome version: Mus_musculus.GRCm39
    - Folders: 2_fastqc + 3_reference -> 4_aligned

# Install libraries

```{r}
#| label: Install libraries
#| include: false

if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install("DESeq2")
BiocManager::install("EnhancedVolcano")
BiocManager::install("tximport")
BiocManager::install("TxDb.Mmusculus.UCSC.mm10.knownGene")
BiocManager::install("biomaRt")
BiocManager::install("sva")
BiocManager::install("msigdbr", force = TRUE)
```

# Import libraries

```{r}
#| label: Load libraries
#| include: false

library(ggplot2)
library(reshape2)
library(tximport)
library(readr)
library(biomaRt)
library(tidyverse)
library(dplyr)
library(DESeq2)
library(EnhancedVolcano)
library(openxlsx)
library(org.Mm.eg.db)
library(AnnotationDbi)
library(tibble)
```

# Functions

## Aggregate by symbol names

```{r}
aggregate_by_rownames <- function(mat) { # mat = txi matrix
  # Convert to dataframe and move rownames to a column
  df <- as.data.frame(mat) %>%
    rownames_to_column(var = "Gene") %>%
    group_by(Gene) %>%
    summarise(across(everything(), ~sum(as.numeric(.), na.rm = TRUE)), .groups = "drop") %>%
    column_to_rownames(var = "Gene")  # Convert back to rownames
  
  return(as.matrix(df))  # Ensure it remains a matrix
}
```

## Top 20 PCA loadings

```{r}
prepare_top_loadings <- function(loadings_vector, top_n = 20) { # loadings_vector = e.g. dds.vsd.pca$rotation[, "PC1"]
  
  top_genes <- head(
    loadings_vector[order(abs(loadings_vector), decreasing = TRUE)],
    top_n
  )
  
  df <- data.frame(
    gene = names(top_genes),
    loading = as.numeric(top_genes)
  )
  
  df$gene <- factor(df$gene, levels = df$gene[order(df$loading)])
  
  return(df)
}
```

# Create raw count matrix

## Obtain raw counts

```{r}
#| warning: false
# Set directory
kallisto_dir <- "../data/4_aligned"

# List all abundance files
samples <- list.files(kallisto_dir, full.names = TRUE)
files <- file.path(samples, "abundance.tsv")
names(files) <- basename(samples)

# Download up to date gene mapping
ensembl <- useEnsembl(biomart = "ensembl", dataset = "mmusculus_gene_ensembl")
tx2gene <- getBM(attributes = c("ensembl_transcript_id", "ensembl_gene_id"),
                       mart = ensembl)
# Save the new mapping
dir.create("./Supplement")
write_tsv(tx2gene, "./Supplement/tx2gene_mouse.tsv", col_names = FALSE)

# Reload it
tx2gene <- read_tsv("./Supplement/tx2gene_mouse.tsv", col_names = c("TXNAME", "GENEID"))
tx2gene$TXNAME <- sub("\\..*", "", tx2gene$TXNAME)

# Import Kallisto quantifications (RAW)----------------------------------------
txi.raw <- tximport(files, type = "kallisto", tx2gene = tx2gene, 
                countsFromAbundance = "no", 
                ignoreTxVersion = TRUE)
```

## Translate Ensembl ID to gene symbols

```{r}
# Get gene symbols
refseq_ids <- as.list(rownames(txi.raw$counts))
refseq_ids <- unlist(refseq_ids)

mouse_ensembl <- useEnsembl(biomart = "ensembl", dataset = "mmusculus_gene_ensembl")

attributes <- c("ensembl_gene_id", "external_gene_name", "gene_biotype")

gene_symbols <- getBM(attributes = attributes,
                      filters = "ensembl_gene_id",
                      values = refseq_ids,
                      mart = mouse_ensembl)

txi.raw$counts <- txi.raw$counts[rownames(txi.raw$counts) %in% gene_symbols$ensembl_gene_id, ]
txi.raw$abundance <- txi.raw$abundance[rownames(txi.raw$abundance) %in% gene_symbols$ensembl_gene_id, ]
txi.raw$length <- txi.raw$length[rownames(txi.raw$length) %in% gene_symbols$ensembl_gene_id, ]

# Add gene symbols
txi.raw$counts <- cbind(symbol = gene_symbols$external_gene_name[match(rownames(txi.raw$counts), gene_symbols$ensembl_gene_id)], 
                          id = rownames(txi.raw$counts), 
                          txi.raw$counts)

txi.raw$abundance <- cbind(symbol = gene_symbols$external_gene_name[match(rownames(txi.raw$abundance), gene_symbols$ensembl_gene_id)], 
                             id = rownames(txi.raw$abundance), 
                             txi.raw$abundance)

txi.raw$length <- cbind(symbol = gene_symbols$external_gene_name[match(rownames(txi.raw$length), gene_symbols$ensembl_gene_id)], 
                          id = rownames(txi.raw$length), 
                          txi.raw$length)

# Remove 'id' column
txi.raw$counts <- txi.raw$counts[, colnames(txi.raw$counts) != "id"]
txi.raw$abundance <- txi.raw$abundance[, colnames(txi.raw$abundance) != "id"]
txi.raw$length <- txi.raw$length[, colnames(txi.raw$length) != "id"]

# Set rownames (symbol)
rownames(txi.raw$counts) <- txi.raw$counts[, "symbol"]
rownames(txi.raw$abundance) <- txi.raw$abundance[, "symbol"]
rownames(txi.raw$length) <- txi.raw$length[, "symbol"]

txi.raw$counts <- txi.raw$counts[, colnames(txi.raw$counts) != "symbol"]
txi.raw$abundance <- txi.raw$abundance[, colnames(txi.raw$abundance) != "symbol"]
txi.raw$length <- txi.raw$length[, colnames(txi.raw$length) != "symbol"]

# Aggregate counts, abundance, and length
txi.raw$counts <- aggregate_by_rownames(txi.raw$counts)
txi.raw$abundance <- aggregate_by_rownames(txi.raw$abundance)
txi.raw$length <- aggregate_by_rownames(txi.raw$length)
```

## Filter samples

```{r}
# Filter unwanted samples
unwanted_samples <- c("YC3", "X2") # Outlier based on PCA and heatmap

txi.raw$counts <- txi.raw$counts[, !colnames(txi.raw$counts) %in% unwanted_samples]
txi.raw$abundance <- txi.raw$abundance[, !colnames(txi.raw$abundance) %in% unwanted_samples]
txi.raw$length <- txi.raw$length[, !colnames(txi.raw$length) %in% unwanted_samples]
```

## Filter genes

```{r}
# Filter unwanted genes 
keep_rows <- !grepl("^(Gm|Rpl)", rownames(txi.raw$counts))

# Apply the filter to all txi components
txi.raw$counts    <- txi.raw$counts[keep_rows, ]
txi.raw$abundance <- txi.raw$abundance[keep_rows, ]
txi.raw$length    <- txi.raw$length[keep_rows, ]
```

# Load metadata

```{r}
colData <- readxl::read_excel("./Supplement/metadata.xlsx")

# Filter samples
colData <- colData[-c(3, 10), ]

# Add rownames
rownames(colData) <- colData$sample

# Reorder to match txi matrix colnames
colData <- colData[colnames(txi.raw$counts), ]
```

# Create DESeq2 object

```{r}
dds.object <- DESeqDataSetFromTximport(txi.raw, colData, ~group)

# Filter out low read counts
smallestGroupSize <- 3
keep <- rowSums(counts(dds.object) >= 10) >= smallestGroupSize
dds.object <- dds.object[keep,]
```

# Check statistics

## Lib sizes

```{r}
lib_size <- data.frame(
  sample = colnames(counts(dds.object)),
  lib_size = colSums(counts(dds.object)) / 1e6
)

lib_size.barplot <- ggplot(lib_size, aes(x = sample, y = lib_size)) +
  geom_col(fill = "#FF4747") +
  labs(y = "Million reads") +
  labs(x = NULL) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

lib_size.barplot
```

```{r}
dir.create("./EDA")
ggsave("./EDA/library_sizes.png", lib_size.barplot, width = 8, height = 4, bg = 'white')
```

## Detected genes

```{r}
detected_genes <- data.frame(
  sample = names(colSums(counts(dds.object) > 10)),
  genes = as.numeric(colSums(counts(dds.object) > 10))
)

detected_genes.barplot <- ggplot(detected_genes, aes(x = sample, y = genes)) +
  geom_col(fill = "#FF4747") +
  labs(y = "Genes with >10 counts") +
  labs(x = NULL) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

detected_genes.barplot
```

```{r}
ggsave("./EDA/detected_genes.png", detected_genes.barplot, width = 8, height = 4, bg = 'white')
```

## Count distribution

```{r}
log_counts.long <- as.data.frame(counts(dds.object)) |>
  rownames_to_column("gene") |>
  pivot_longer(
    -gene,
    names_to = "sample",
    values_to = "count"
  ) |>
  mutate(log_count = log10(count + 1))

count_dist.boxplot <- ggplot(log_counts.long, aes(x = sample, y = log_count)) +
  geom_boxplot(
    outlier.shape = NA,
    fill = "#FF4747"
  ) +
  labs(
    x = NULL,
    y = "log10(counts + 1)"
  ) +
  theme_bw() +
  ylim(0, 5) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1)
  )

count_dist.boxplot
```

```{r}
ggsave("./EDA/count_distribution.png", count_dist.boxplot, width = 8, height = 4, bg = 'white')
```

# Run DESeq2

```{r}
# Run DESeq2 
dds <- DESeq(dds.object)

# Save object
saveRDS(dds, file = "./EDA/dds.rds")
```

# Check statistics

## Density plot

```{r}
# Transform dds (variance stabilization for EDA)
vsd <- vst(dds, blind = FALSE)

# Get matrix
vsd_mat <- assay(vsd)

# Create df for density plot
density.df <- melt(vsd_mat)
colnames(density.df) <- c("gene", "sample", "expression")

density.df$group <- colData(vsd)$group[density.df$sample]

density_plot <- ggplot(density.df, aes(x = expression, group = sample, color = group)) +
  geom_density(alpha = 0.3, linewidth = 0.6) +
  theme_bw() +
  labs(
    title = "Sample-specific density of VST expression",
    x = "VST expression",
    y = "Density"
  )

density_plot
```

```{r}
ggsave("./EDA/density_plot.png", density_plot, width = 8, height = 4, bg = 'white')
```

## Dispersion plot

```{r}
plotDispEsts(dds)
```

# PCA (with variance stabilization)

```{r}
# Run PCA
dds.vsd.pca <- prcomp(t(vsd_mat), scale. = FALSE)

# Create a data frame with PCA results
dds.vsd.pca.df <- as.data.frame(dds.vsd.pca$x)
dds.vsd.pca.df$sample <- rownames(dds.vsd.pca.df)  # Add sample names

# Add metadata
dds.vsd.pca.df <- dds.vsd.pca.df %>%
  dplyr::left_join(colData %>% dplyr::select(sample, group, sex), by = c("sample" = "sample")) # adjust according to metadata columns
  
colnames(dds.vsd.pca.df)[18] <- "Group"

# Plot PCA
dds.vsd.pca.plot <- ggplot(dds.vsd.pca.df, aes(x = PC1, y = PC2, label = sample, color = Group, fill = Group)) +
  geom_point(shape = 21, size = 6, stroke = 0.5) + 
  
  geom_text(vjust = -2, show.legend = FALSE, color = "grey40", size = 5) +
  theme_minimal() +
  
  scale_color_manual(values = c("#9B111E", "#00072D", "grey60")) + 
  scale_fill_manual(values = c("#FF4747", "steelblue", "grey")) + 
  
  theme(
    text = element_text(size = 24),        
    axis.title = element_text(size = 24),    
    axis.text = element_text(size = 22),    
    legend.text = element_text(size = 22),  
    legend.title = element_text(size = 24),
    legend.key.spacing.y = unit(0.2, "cm")
  )

dds.vsd.pca.plot

# YC7 and X2 might be an outlier -> filter above
```

```{r}
ggsave("./EDA/PCA_filtered.png", dds.vsd.pca.plot, width = 12, height = 12, bg = 'white', dpi = 300)
```

## Examine top variables

```{r}
# Obtain loadings
loadings <- dds.vsd.pca$rotation

# Top genes positively and negatively contributing to PC1
top_PC1_df <- prepare_top_loadings(loadings[, "PC1"], top_n = 20)

# Barplot
top_PC1.barplot <- ggplot(top_PC1_df, aes(x = gene, y = loading, fill = loading > 0)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(values = c("steelblue", "#FF4747"), guide = FALSE) +
  labs(title = "Top PC1 genes (Aged vs Young)", x = "Gene", y = "Loading") +
  theme_minimal()

top_PC1.barplot

# Top genes positively and negatively contributing to PC2
top_PC2_df <- prepare_top_loadings(loadings[, "PC2"], top_n = 20)

# Barplot
top_PC2.barplot <- ggplot(top_PC2_df, aes(x = gene, y = loading, fill = loading > 0)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(values = c("steelblue", "#FF4747"), guide = FALSE) +
  labs(title = "Top PC2 genes (Aged vs Young)", x = "Gene", y = "Loading") +
  theme_minimal()

top_PC2.barplot
```

```{r}
ggsave("./EDA/PC1_barplot_filtered.png", top_PC1.barplot, width = 5, height = 5, bg = 'white')
ggsave("./EDA/PC2_barplot_filtered.png", top_PC2.barplot, width = 5, height = 5, bg = 'white')
```

# Heatmap

## Top variable genes heatmap

```{r}
vars <- apply(assay(vsd), 1, var)
top_genes <- head(order(vars, decreasing = TRUE), 500)

pheatmap::pheatmap(
  assay(vsd)[top_genes, ],
  scale = "row",
  show_rownames = FALSE
)

pheatmap::pheatmap(
  assay(vsd)[top_genes, ],
  scale = "row",
  show_rownames = FALSE,
  filename = "EDA/vst_top_variable_genes_heatmap.png",
  width = 10,
  height = 10
)
```

## Correlation heatmap

```{r}
pheatmap::pheatmap(
  cor(assay(vsd))
  )

pheatmap::pheatmap(
  cor(assay(vsd)),
  filename = "EDA/vst_correlation_heatmap.png",
  width = 10,
  height = 10
  )
```
