---
title: "Ageing + X bulk mRNAseq (DEGs)"
format: html
author: Tamas Lakat
---

# Experiment

The dataset contains bulk mRNA-sequencing data obtained from mouse heart samples (Young Control (YC; n=4), Aged Control (AC; n=4), Aged Treated (X; n=10))

# Preprocessing pipeline

1.  **Trimming + QC**
    - Software: fastp
    - Folders: 0_raw ->  1_trimmed
2.  **QC**
    - Software: fastqc, multiqc
    - Folders: 1_trimmed -> 2_fastqc
3.  **Pseudoalignment**
    - Software: Kallisto
    - Reference transcriptome version: Mus_musculus.GRCm39
    - Folders: 2_fastqc + 3_reference -> 4_aligned

# Install libraries

```{r}
#| label: Install libraries
#| include: false

if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install("DESeq2")
BiocManager::install("EnhancedVolcano")
BiocManager::install("tximport")
BiocManager::install("TxDb.Mmusculus.UCSC.mm10.knownGene")
BiocManager::install("biomaRt")
BiocManager::install("sva")
BiocManager::install("msigdbr", force = TRUE)
```

# Import libraries

```{r}
#| label: Load libraries
#| include: false

library(ggplot2)
library(tximport)
library(readr)
library(biomaRt)
library(tidyverse)
library(dplyr)
library(DESeq2)
library(EnhancedVolcano)
library(openxlsx)
library(org.Mm.eg.db)
library(AnnotationDbi)
library(ComplexHeatmap)
library(colorRamp2)
library(ragg)
library(clusterProfiler)
```

# Functions

## Create color vectors for volcano plot

```{r}
color_vector <- function(df, log2FC = 1.5, alpha = 0.05) {
  
  keyvals <- rep("grey", nrow(df))
  names(keyvals) <- rep("NS", nrow(df))  # ← rep() to give every element its own "NS" name
  
  valid <- !is.na(df$log2FoldChange) & !is.na(df$padj)
  
  up <- valid & df$log2FoldChange > log2FC & df$padj < alpha
  keyvals[up] <- "#FF4747"
  names(keyvals)[up] <- "Upregulated"
  
  down <- valid & df$log2FoldChange < -log2FC & df$padj < alpha
  keyvals[down] <- "steelblue"
  names(keyvals)[down] <- "Downregulated"
  
  return(keyvals)
}
```

## Check significance

```{r}
is_sig <- function(padj, lfc) {
  (padj < 0.05 | padj == 0) & abs(lfc) > 1.5 & abs(lfc) < 10
}
```

# Load data

```{r}
dds <- readRDS("./EDA/dds.Rds")
```

# Pairwise comparison

```{r}
# Pairwise comparison
aged_vs_young <- results(dds, contrast=c("group", "Aged Control", "Young Control"))
x_vs_aged <- results(dds, contrast=c("group", "Aged Treated", "Aged Control"))
x_vs_young <- results(dds, contrast=c("group", "Aged Treated", "Young Control"))

# Convert to df
aged_vs_young.df <- as.data.frame(aged_vs_young)
aged_vs_young.df <- aged_vs_young.df[!is.na(aged_vs_young.df$padj), ]

x_vs_aged.df <- as.data.frame(x_vs_aged)
x_vs_aged.df <- x_vs_aged.df[!is.na(x_vs_aged.df$padj), ]

x_vs_young.df <- as.data.frame(x_vs_young)
x_vs_young.df <- x_vs_young.df[!is.na(x_vs_young.df$padj), ]
```

```{r}
# Save results
dir.create("./DEGs")
saveRDS(aged_vs_young, file = "./DEGs/aged_vs_young.rds")
saveRDS(x_vs_aged, file = "./DEGs/x_vs_aged.rds")
```

# Check statistics

## FC distribution

```{r}
aged_vs_young.lfc.df <- data.frame(
  log2FC = aged_vs_young$log2FoldChange
)

aged_vs_young.lfc.hist <- ggplot(aged_vs_young.lfc.df, aes(x = log2FC)) +
  geom_histogram(
    bins = 50,
    fill = "#FF4747",
    color = "black"
  ) +
  labs(
    title = "Log2 Fold Changes Distribution (Aged vs Young)",
    x = "Log2 Fold Change",
    y = "Gene count"
  ) +
  theme_bw()

aged_vs_young.lfc.hist

x_vs_aged.lfc.df <- data.frame(
  log2FC = x_vs_aged$log2FoldChange
)

x_vs_aged.lfc.hist <- ggplot(x_vs_aged.lfc.df, aes(x = log2FC)) +
  geom_histogram(
    bins = 50,
    fill = "#FF4747",
    color = "black"
  ) +
  labs(
    title = "Log2 Fold Changes Distribution (x vs Aged)",
    x = "Log2 Fold Change",
    y = "Gene count"
  ) +
  theme_bw()

x_vs_aged.lfc.hist
```

```{r}
ggsave("./DEGs/aged_vs_young_lfc_hist.png", aged_vs_young.lfc.hist, width = 8, height = 4, bg = 'white')
ggsave("./DEGs/x_vs_aged_lfc_hist.png", x_vs_aged.lfc.hist, width = 8, height = 4, bg = 'white')
```

## Padj distribution

```{r}
aged_vs_young.padj.df <- data.frame(
  padj = aged_vs_young$padj
)

aged_vs_young.padj.hist <- ggplot(aged_vs_young.padj.df, aes(x = padj)) +
  geom_histogram(
    bins = 50,
    fill = "#FF4747",
    color = "black"
  ) +
  labs(
    title = "Adjusted P Distribution (Aged vs Young)",
    x = "padj",
    y = "Gene count"
  ) +
  theme_bw()

aged_vs_young.padj.hist

x_vs_aged.padj.df <- data.frame(
  padj = x_vs_aged$padj
)

x_vs_aged.padj.hist <- ggplot(x_vs_aged.padj.df, aes(x = padj)) +
  geom_histogram(
    bins = 50,
    fill = "#FF4747",
    color = "black"
  ) +
  labs(
    title = "Adjusted P Distribution (x vs Aged)",
    x = "padj",
    y = "Gene count"
  ) +
  theme_bw()

x_vs_aged.padj.hist
```

```{r}
ggsave("./DEGs/aged_vs_young_padj_hist.png", aged_vs_young.padj.hist, width = 8, height = 4, bg = 'white')
ggsave("./DEGs/x_vs_aged_padj_hist.png", x_vs_aged.padj.hist, width = 8, height = 4, bg = 'white')
```

# Volcano plot

```{r}
# Define the significance threshold
alpha <- 0.05
log2FC <- 1.5

# Create a custom color vector based on log2 fold change and p-value (Check Functions!)
aged_vs_young.keyvals <- color_vector(aged_vs_young.df)
x_vs_aged.keyvals <- color_vector(x_vs_aged.df)
x_vs_young.keyvals <- color_vector(x_vs_young.df)
  
# Create the volcano plot
# Aged vs young
aged_vs_young.volcano <- EnhancedVolcano(aged_vs_young.df,
                                            lab = rownames(aged_vs_young.df),
                                            x = "log2FoldChange",
                                            y = "padj",
                                            pCutoff = alpha,
                                            FCcutoff = log2FC,
                                            pointSize = 3.0,
                                            labSize = 3.0,
                                            xlim = c(-8, 8),
                                            ylim = c(0, 10),
                                            title = "Aged vs Young",
                                            subtitle = NULL,
                                            caption = NULL,
                                            gridlines.major = FALSE,
                                            gridlines.minor = FALSE,
                                            colCustom = aged_vs_young.keyvals
                                          )



aged_vs_young.volcano

# x vs aged
x_vs_aged.volcano <- EnhancedVolcano(x_vs_aged.df,
                                            lab = rownames(x_vs_aged.df),
                                            x = "log2FoldChange",
                                            y = "padj",
                                            pCutoff = alpha,
                                            FCcutoff = log2FC,
                                            pointSize = 3.0,
                                            labSize = 3.0,
                                            xlim = c(-8, 8),
                                            ylim = c(0, 10),
                                            title = "x vs Aged",
                                            subtitle = NULL,
                                            caption = NULL,
                                            gridlines.major = FALSE,
                                            gridlines.minor = FALSE,
                                            colCustom = x_vs_aged.keyvals
                                          )


x_vs_aged.volcano

# x vs young
x_vs_young.volcano <- EnhancedVolcano(x_vs_young.df,
                                            lab = rownames(x_vs_young.df),
                                            x = "log2FoldChange",
                                            y = "padj",
                                            pCutoff = alpha,
                                            FCcutoff = log2FC,
                                            pointSize = 3.0,
                                            labSize = 3.0,
                                            xlim = c(-8, 8),
                                            ylim = c(0, 10),
                                            title = "x vs Young",
                                            subtitle = NULL,
                                            caption = NULL,
                                            gridlines.major = FALSE,
                                            gridlines.minor = FALSE,
                                            colCustom = x_vs_young.keyvals
                                          )


x_vs_young.volcano
```

```{r}
ggsave("./DEGs/volcano_aged_vs_young.png", aged_vs_young.volcano, width = 8, height = 8, bg = 'white', dpi = 300)
ggsave("./DEGs/volcano_x_vs_aged.png", x_vs_aged.volcano, width = 8, height = 8, bg = 'white', dpi = 300)
ggsave("./DEGs/volcano_x_vs_young.png", x_vs_young.volcano, width = 8, height = 8, bg = 'white', dpi = 300)
```

# Export DEGs

## Create DEGs df (sorted)

```{r}
# Aged vs young
aged_vs_young.df.significant <- aged_vs_young.df %>%
  tibble::rownames_to_column("symbol") %>%
  filter(!is.na(padj), !is.na(log2FoldChange)) %>%
  filter(padj < 0.05) %>%
  arrange(desc(log2FoldChange))

rownames(aged_vs_young.df.significant) <- aged_vs_young.df.significant$symbol

# x vs aged
x_vs_aged.df.significant <- x_vs_aged.df %>%
  tibble::rownames_to_column("symbol") %>%
  filter(!is.na(padj), !is.na(log2FoldChange)) %>%
  filter(padj < 0.05) %>%
  arrange(desc(log2FoldChange))

rownames(x_vs_aged.df.significant) <- x_vs_aged.df.significant$symbol


```

## Export to excel

```{r}
# Create a new Excel workbook
wb <- createWorkbook()

# List of data frames and corresponding sheet names
df_list <- list(
  "Aged_vs_Young" = aged_vs_young.df.significant,
  "x_vs_Aged" = x_vs_aged.df.significant
)

# Loop through the list and add each data frame as a new sheet
for (sheet_name in names(df_list)) {
  addWorksheet(wb, sheet_name)  # Create a new sheet
  writeData(wb, sheet = sheet_name, df_list[[sheet_name]])  # Write data
}

# Save the workbook as an Excel file
saveWorkbook(wb, "./DEGs/DEGs_filtered.xlsx", overwrite = TRUE)
```

# Barplots

```{r}
# Aged_vs_young
## Filter top 20 DEGs
aged_vs_young.top_pos <- head(aged_vs_young.df.significant, 10)
aged_vs_young.top_neg <- tail(aged_vs_young.df.significant, 10)
aged_vs_young.top_20 <- bind_rows(aged_vs_young.top_pos, aged_vs_young.top_neg)

## Create barplot
aged_vs_young.top_20.barplot <- ggplot(aged_vs_young.top_20, aes(x = log2FoldChange, y=reorder(rownames(aged_vs_young.top_20), log2FoldChange), fill = padj)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_gradient(low = "#FF4747", high = "azure", na.value = "azure") +
  labs(
    x = "log2 Fold Change",
    y = "Gene",
    fill = "Adjusted p-value",
    title = "Top 20 DEGs (Aged vs Young)"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 10),  # Adjust y-axis gene label size
    axis.text.x = element_text(size = 10),  # Adjust x-axis text size
    legend.position = "right"
  )
  
aged_vs_young.top_20.barplot

## x vs aged
x_vs_aged.top_pos <- head(x_vs_aged.df.significant, 10)
x_vs_aged.top_neg <- tail(x_vs_aged.df.significant, 10)
x_vs_aged.top_20 <- bind_rows(x_vs_aged.top_pos, x_vs_aged.top_neg)

## Create barplot
x_vs_aged.top_20.barplot <- ggplot(x_vs_aged.top_20, aes(x = log2FoldChange, y=reorder(rownames(x_vs_aged.top_20), log2FoldChange), fill = padj)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_gradient(low = "#FF4747", high = "azure", na.value = "azure") +
  labs(
    x = "log2 Fold Change",
    y = "Gene",
    fill = "Adjusted p-value",
    title = "Top 20 DEGs (x vs Aged)"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 10),  # Adjust y-axis gene label size
    axis.text.x = element_text(size = 10),  # Adjust x-axis text size
    legend.position = "right"
  )
  
x_vs_aged.top_20.barplot
```

```{r}
ggsave("./DEGs/barplot_top20_aged_vs_young.png", aged_vs_young.top_20.barplot, width = 8, height = 8, bg = 'white')
ggsave("./DEGs/barplot_top20_x_vs_aged.png", x_vs_aged.top_20.barplot, width = 8, height = 8, bg = 'white')
```

# Scatterplot (Pearson correlation)

```{r}
# Get the union of significant genes from both datasets
union_genes <- union(aged_vs_young.df.significant$symbol, x_vs_aged.df.significant$symbol)

# Only keep union genes that exist in each df
aged_genes   <- intersect(union_genes, rownames(aged_vs_young.df))
x_genes    <- intersect(union_genes, rownames(x_vs_aged.df))
common_genes <- intersect(aged_genes, x_genes)  # genes present in both

# Subset from filtered dataframes, not raw DESeqResults
aged_vs_young.df.union <- aged_vs_young.df[common_genes, ]
x_vs_aged.df.union   <- x_vs_aged.df[common_genes, ]

# Now merge — same genes guaranteed, so suffix approach is safe
scatter_data.merged <- merge(
  aged_vs_young.df.union |> tibble::rownames_to_column("gene"),
  x_vs_aged.df.union   |> tibble::rownames_to_column("gene"),
  by = "gene", suffixes = c("_aged", "_x")
)

# Add a column to indicate significance
scatter_data.merged$significance <- ifelse(
  is_sig(scatter_data.merged$padj_aged, scatter_data.merged$log2FoldChange_aged) &
    is_sig(scatter_data.merged$padj_x, scatter_data.merged$log2FoldChange_x),
  "Both",
  ifelse(
    is_sig(scatter_data.merged$padj_aged, scatter_data.merged$log2FoldChange_aged),
    "Aged",
    ifelse(
      is_sig(scatter_data.merged$padj_x, scatter_data.merged$log2FoldChange_x),
      "x",
      "None"
    )
  )
)

scatter_data.merged <- scatter_data.merged[scatter_data.merged$significance!="None", ]

scatter_data.merged[is.na(scatter_data.merged)] <- 0

# Scatter plot with Pearson correlation
scatterplot.pearson <- ggplot(scatter_data.merged, aes(x = log2FoldChange_aged, y = log2FoldChange_x, color = significance)) +
  geom_point(size = 3.5, alpha = 0.6) +  
  geom_smooth(method = "lm", color = "red", linetype = "dashed") +  
  
  annotate("text", x = max(scatter_data.merged$log2FoldChange_aged) - 5, 
           y = max(scatter_data.merged$log2FoldChange_x) - 3, 
           label = paste("Pearson R =", round(cor(scatter_data.merged$log2FoldChange_aged, 
                                                 scatter_data.merged$log2FoldChange_x), 2),
                         "\np-value ", format.pval(cor.test(scatter_data.merged$log2FoldChange_aged, 
                                                            scatter_data.merged$log2FoldChange_x)$p.value, 
                                                   digits = 3)),
           size = 7, color = "black", hjust = 0) +  
  
  geom_vline(xintercept = c(-1.5, 1.5), linetype = "dashed", color = "darkgrey", linewidth = 0.4) +  
  geom_hline(yintercept = c(-1.5, 1.5), linetype = "dashed", color = "darkgrey", linewidth = 0.4) +  
  geom_vline(xintercept = c(-0, 0), color = "black", linewidth = 0.2) +  
  geom_hline(yintercept = c(-0, 0), color = "black", linewidth = 0.2) +  
  labs(title = "",
       x = "log2FoldChange (Aged vs Young)",
       y = "log2FoldChange (x vs Aged)") +
  scale_color_manual(values = c("Aged" = "lightblue", "x" = "steelblue", "Both" = "#FF4747", "None" = "gray")) +  
  theme_minimal() +
  theme(
    plot.title = element_text(size = 22, face = "bold"),  
    axis.line = element_line(color = "black", linewidth = 0.8),  
    axis.text = element_text(size = 22),  
    axis.title = element_text(size = 24),  
    legend.text = element_text(size = 22),
    legend.key.spacing.y = unit(0.2, "cm")
  ) +
  guides(color = guide_legend(title = NULL))  


scatterplot.pearson
```

```{r}
ggsave("./DEGs/scatterplot_pearson.png", scatterplot.pearson, width = 12, height = 12, bg = 'white', dpi = 300)
```

## Heatmap of correlating genes

```{r}
# Transform dds (variance stabilization)
vsd <- vst(dds, blind = FALSE)

# Extract significant genes labeled as "Both"
both <- scatter_data.merged$gene[scatter_data.merged$significance == "Both"]
both <- intersect(both, rownames(vsd))

# Subset count matrix for these genes
both.matrix <- assay(vsd)[both, ]

# Ensure no missing genes (remove NA rows)
both.matrix <- both.matrix[complete.cases(both.matrix), ]

# Z-score normalization
both.matrix.z <- t(scale(t(both.matrix)))

# Define the desired sample order
desired_order <- c("YC1", "YC2", "YC3", "YC4",
                   "AC1", "AC2", "AC3", "AC4",
                   "X1", "X2", "X3", "X4", "X5", "X6", "X7", "X8", "X9", "X10")

# Ensure only available samples are used
desired_order <- intersect(desired_order, colnames(both.matrix.z))

# Reorder the columns in the matrix based on the specified order
both.matrix.z <- both.matrix.z[, desired_order]

# Create heatmap with your preferred styling
both.matrix.z.heatmap <- Heatmap(
  both.matrix.z, 
  cluster_rows = TRUE, 
  cluster_columns = FALSE, 
  col = colorRamp2(c(-2, 0, 2), c("steelblue", "white", "#FF4747")),  # Color scheme
  show_row_dend = FALSE,    # Hide row dendrogram
  show_column_dend = FALSE, # Hide column dendrogram
  show_heatmap_legend = TRUE,
  heatmap_legend_param = list(
    title = "z-score",  
    title_gp = gpar(fontsize = 14),  
    labels_gp = gpar(fontsize = 12)  
  ),
  rect_gp = gpar(col = "white", lwd = 1),  # Add white grid
  column_names_gp = gpar(fontsize = 10),  # Adjust column names font size
  row_names_gp = gpar(fontsize = 10),  # Adjust row names font size
  row_names_max_width = unit(35, "cm")  # Set max row name width
)

both.matrix.z.heatmap
```

```{r}
tryCatch({
    # Open the device
    png("DEGs/correlating_genes_heatmap.png", 
        width = 1600,
        height = 3600,
        res = 300)
    
    # Calculate and Draw
    print("Drawing heatmap...")
    ComplexHeatmap::draw(both.matrix.z.heatmap)
    
    # Force the device to save
    dev.off()
    print("File saved successfully!")
    
}, error = function(e) {
    print(paste("Error caught:", e$message))
    if(!is.null(dev.list())) dev.off()
})

```

# Protein-protein interaction

## Export gene list

```{r}
x_vs_aged.significant.gene_list <- x_vs_aged.df.significant %>%
  filter(abs(log2FoldChange) > 1.5) %>%
  dplyr::select(symbol)

dir.create("./PPI")
write.table(x_vs_aged.significant.gene_list, file = "./PPI/signif_gene_list_x_vs_aged.txt", sep = "\t", row.names = FALSE)
```

## Import STRING network dataframe

```{r}
# Read STRING analysis results
x_vs_aged.string <- read.csv("./PPI/x_vs_aged_10plus_degree_proteins.csv", sep = ",")

x_vs_aged.string.degree <- as.data.frame(x_vs_aged.string) %>%
  dplyr::select(display.name, Degree) %>%
  na.omit()

x_vs_aged.topdegree_genes <- x_vs_aged.string.degree$display.name[x_vs_aged.string.degree$Degree>9]
```

## Heatmap of top degree genes

```{r}
# Extract valid genes
x_vs_aged.topdegree_genes <- intersect(x_vs_aged.topdegree_genes, rownames(vsd))

# Subset count matrix for these genes
x_vs_aged.topdegree_genes.matrix <- assay(vsd)[x_vs_aged.topdegree_genes, ]

# Ensure no missing genes (remove NA rows)
x_vs_aged.topdegree_genes.matrix <- x_vs_aged.topdegree_genes.matrix[complete.cases(x_vs_aged.topdegree_genes.matrix), ]

# Z-score normalization
x_vs_aged.topdegree_genes.matrix.z <- t(scale(t(x_vs_aged.topdegree_genes.matrix)))

# Define the desired sample order
desired_order <- c("YC1", "YC2", "YC3", "YC4",
                   "AC1", "AC2", "AC3", "AC4",
                   "X1", "X2", "X3", "X4", "X5", "X6", "X7", "X8", "X9", "X10")

# Ensure only available samples are used
desired_order <- intersect(desired_order, colnames(x_vs_aged.topdegree_genes.matrix.z))

# Reorder the columns in the matrix based on the specified order
x_vs_aged.topdegree_genes.matrix.z <- x_vs_aged.topdegree_genes.matrix.z[, desired_order]

# Create heatmap with your preferred styling
x_vs_aged.topdegree_genes.matrix.z.heatmap <- Heatmap(
  x_vs_aged.topdegree_genes.matrix.z, 
  cluster_rows = TRUE, 
  cluster_columns = FALSE, 
  col = colorRamp2(c(-2, 0, 2), c("steelblue", "white", "#FF4747")),  # Color scheme
  show_row_dend = FALSE,    # Hide row dendrogram
  show_column_dend = FALSE, # Hide column dendrogram
  show_heatmap_legend = TRUE,
  heatmap_legend_param = list(
    title = "z-score",  
    title_gp = gpar(fontsize = 14),  
    labels_gp = gpar(fontsize = 12)  
  ),
  rect_gp = gpar(col = "white", lwd = 1),  # Add white grid
  column_names_gp = gpar(fontsize = 10),  # Adjust column names font size
  row_names_gp = gpar(fontsize = 10),  # Adjust row names font size
  row_names_max_width = unit(35, "cm")  # Set max row name width
)

x_vs_aged.topdegree_genes.matrix.z.heatmap
```

```{r}
tryCatch({
    # Open the device
    png("PPI/x_vs_aged_topdegree_genes_heatmap.png", 
        width = 2000,
        height = 2000,
        res = 300)
    
    # Calculate and Draw
    print("Drawing heatmap...")
    ComplexHeatmap::draw(x_vs_aged.topdegree_genes.matrix.z.heatmap)
    
    # Force the device to save
    dev.off()
    print("File saved successfully!")
    
}, error = function(e) {
    print(paste("Error caught:", e$message))
    if(!is.null(dev.list())) dev.off()
})

```

## ORA of top degree genes

```{r}
# Set gene set size thresholds
minGSSize = 15
maxGSSize = 300

# Obtain go results
x_vs_aged_topdegree_genes.go <- enrichGO(
  gene = x_vs_aged.topdegree_genes,
  OrgDb = org.Mm.eg.db,
  ont = 'ALL', # Choose ALL, BP, MF, CC
  pvalueCutoff = 0.05,
  minGSSize = minGSSize,
  maxGSSize = maxGSSize,
  keyType = "SYMBOL")

# Visualize GO results
x_vs_aged_topdegree_genes.go.dotplot <- dotplot(x_vs_aged_topdegree_genes.go) + ggtitle("GO: x vs Aged Topdegree Genes")

x_vs_aged_topdegree_genes.go.dotplot
```

```{r}
ggsave("./PPI/x_vs_aged_topdegree_genes_ora_go.png", x_vs_aged_topdegree_genes.go.dotplot, width = 8, height = 8, bg = 'white')
```
