---
title: "Ageing + X bulk mRNAseq (Functional enrichment analysis)"
format: html
author: Tamas Lakat
---

# Experiment

The dataset contains bulk mRNA-sequencing data obtained from mouse heart samples (Young Control (YC; n=4), Aged Control (AC; n=4), Aged Treated (X; n=10))

# Preprocessing pipeline

1.  **Trimming + QC**
    - Software: fastp
    - Folders: 0_raw ->  1_trimmed
2.  **QC**
    - Software: fastqc, multiqc
    - Folders: 1_trimmed -> 2_fastqc
3.  **Pseudoalignment**
    - Software: Kallisto
    - Reference transcriptome version: Mus_musculus.GRCm39
    - Folders: 2_fastqc + 3_reference -> 4_aligned

# Install libraries

```{r}
#| label: Install libraries
#| include: false

if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install("clusterProfiler")
BiocManager::install("enrichplot")
BiocManager::install("org.Mm.eg.db")
BiocManager::install("biomaRt")
BiocManager::install("ReactomePA")
```

# Import libraries

```{r}
#| label: Load libraries
#| include: false

library(msigdbr)
library(org.Mm.eg.db)
library(biomaRt)
library(clusterProfiler)
library(readr)
library(tidyverse)
library(dplyr)
library(DESeq2)
library(openxlsx)
library(enrichplot)
library(ReactomePA)
library(DT)
```

# Functions

## Numeric vector for ORA

```{r}
numeric_vector <- function(df, organism) { # df = DESeq2 results as df, organism: mouse = org.Mm.eg.db
  df.significant <- df |>
    dplyr::filter(
      !is.na(padj),
      padj < padj_threshold,
      abs(log2FoldChange) > log2fc_threshold
    )
  
  df.gene_entrezid <- AnnotationDbi::select(
    organism,
    keys = rownames(df.significant),
    columns = "ENTREZID",
    keytype = "SYMBOL"
  )
  
  df.gene_entrezid <- df.gene_entrezid |>
    dplyr::filter(!is.na(ENTREZID))
  
  df.significant$SYMBOL <- rownames(df.significant)
  
  df.mapped <- dplyr::inner_join(
    df.significant,
    df.gene_entrezid,
    by = "SYMBOL"
  )
  
  df.gene_list <- df.mapped$log2FoldChange
  names(df.gene_list) <- df.mapped$ENTREZID
  
  return(df.gene_list)
}
```

## Ranked gene list for GSEA

```{r}
ranked_gene_list <- function(df) { # df = DESeq2 results as df
  df.ranked <- df %>%
    as.data.frame() %>%
    mutate(
      symbol = rownames(.),
      rank_metric = sign(log2FoldChange) * -log10(padj)
    )

  ranked_genes <- df.ranked$rank_metric
  names(ranked_genes) <- df.ranked$symbol

  ranked_genes <- sort(ranked_genes, decreasing = TRUE)

  return(ranked_genes)
}
```

## List top 10 pos and top 10 neg GSEA results

```{r}
# Perform GSEA
gsea_top20 <- function(res.GSEA) { # GSEA results

  res.GSEA.df <- as_tibble(res.GSEA@result) %>%
    mutate(
      direction = ifelse(NES > 0, "positive", "negative"),
      neglog10_padj = -log10(p.adjust),
    )

  top_positive <- res.GSEA.df %>%
    filter(NES > 0) %>%
    arrange(desc(NES)) %>%
    slice_head(n = 10)

  top_negative <- res.GSEA.df %>%
    filter(NES < 0) %>%
    arrange(NES) %>%
    slice_head(n = 10)

  top20 <- bind_rows(top_positive, top_negative) %>%
    mutate(ID = reorder(ID, NES))

  return(top20)
}
```

# Load DEGs

```{r}
aged_vs_young <- readRDS("./DEGs/aged_vs_young.rds")
x_vs_aged <- readRDS("./DEGs/x_vs_aged.rds")

aged_vs_young.df <- as.data.frame(aged_vs_young)
x_vs_aged.df <- as.data.frame(x_vs_aged)
```

# Overrepresentation analysis (ORA)

## Create gene list (numeric vectors for cnet plot, Check Functions) (or try keyType = "SYMBOL" and simply use symbol instead of ENTREZID)

```{r}
log2fc_threshold <- 1.5
padj_threshold <- 0.05

# numeric_vector (df, organism) df = DESeq2 results as df, organism: mouse = org.Mm.eg.db
aged_vs_young.gene_list <- numeric_vector(aged_vs_young.df, org.Mm.eg.db)
x_vs_aged.gene_list <- numeric_vector(x_vs_aged.df, org.Mm.eg.db)

# Overlap
overlapping_genes <- intersect(names(aged_vs_young.gene_list), names(x_vs_aged.gene_list))
```

## KEGG

```{r}
# Obtain KEGG results
aged_vs_young.kegg <- enrichKEGG(
  gene = names(aged_vs_young.gene_list), 
  organism = 'mmu',  # mmu for mouse, hsa for human
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH")

x_vs_aged.kegg <- enrichKEGG(
  gene = names(x_vs_aged.gene_list), 
  organism = 'mmu',  # mmu for mouse, hsa for human
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH")

overlap.kegg <- enrichKEGG(
  gene = overlapping_genes, 
  organism = 'mmu',  # mmu for mouse, hsa for human
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH")

# Visualize KEGG results
aged_vs_young.kegg.dotplot <- dotplot(aged_vs_young.kegg) + ggtitle("KEGG: Aged vs Young")
x_vs_aged.kegg.dotplot <- dotplot(x_vs_aged.kegg) + ggtitle("KEGG: x vs Aged")
overlap.kegg.dotplot <- dotplot(overlap.kegg) + ggtitle("KEGG: Overlap")

aged_vs_young.kegg.dotplot
x_vs_aged.kegg.dotplot
overlap.kegg.dotplot
```

```{r}
dir.create("./ORA")

ggsave("./ORA/aged_vs_young_ora_kegg.png", aged_vs_young.kegg.dotplot, width = 8, height = 8, bg = 'white')
ggsave("./ORA/x_vs_aged_ora_kegg.png", x_vs_aged.kegg.dotplot, width = 8, height = 8, bg = 'white')
ggsave("./ORA/overlap_ora_kegg.png", overlap.kegg.dotplot, width = 8, height = 8, bg = 'white')
```

## GO

```{r}
# Set gene set size thresholds
minGSSize = 15
maxGSSize = 300

# Obtain go results
aged_vs_young.go <- enrichGO(
  gene = names(aged_vs_young.gene_list),
  OrgDb = org.Mm.eg.db,
  ont = 'BP', # Choose ALL, BP, MF, CC
  pvalueCutoff = 0.05,
  minGSSize = minGSSize,
  maxGSSize = maxGSSize)

x_vs_aged.go <- enrichGO(
  gene = names(x_vs_aged.gene_list),
  OrgDb = org.Mm.eg.db,
  ont = 'BP', # Choose ALL, BP, MF, CC
  pvalueCutoff = 0.05,
  minGSSize = minGSSize,
  maxGSSize = maxGSSize)

overlap.go <- enrichGO(
  gene = overlapping_genes,
  OrgDb = org.Mm.eg.db,
  ont = 'BP', # Choose ALL, BP, MF, CC
  pvalueCutoff = 0.05,
  minGSSize = minGSSize,
  maxGSSize = maxGSSize)

# Visualize GO results
aged_vs_young.go.dotplot <- dotplot(aged_vs_young.go) + ggtitle("GOBP: Aged vs Young")
x_vs_aged.go.dotplot <- dotplot(x_vs_aged.go) + ggtitle("GOBP: x vs Aged")
overlap.go.dotplot <- dotplot(overlap.go) + ggtitle("GOBP: Overlap")

aged_vs_young.go.dotplot
x_vs_aged.go.dotplot
overlap.go.dotplot
```

```{r}
ggsave("./ORA/aged_vs_young_ora_gobp.png", aged_vs_young.go.dotplot, width = 8, height = 8, bg = 'white')
ggsave("./ORA/x_vs_aged_ora_gobp.png", x_vs_aged.go.dotplot, width = 8, height = 8, bg = 'white')
ggsave("./ORA/overlap_ora_gobp.png", overlap.go.dotplot, width = 8, height = 8, bg = 'white')
```

### Enrich plot

```{r}
# Obtain emap results
aged_vs_young.go.emap <- pairwise_termsim(aged_vs_young.go)
x_vs_aged.go.emap <- pairwise_termsim(x_vs_aged.go)
overlap.go.emap <- pairwise_termsim(overlap.go)

# Visualize emap
aged_vs_young.go.emapplot <- emapplot(aged_vs_young.go.emap) + ggtitle("GOBP: Aged vs Young")
x_vs_aged.go.emapplot <- emapplot(x_vs_aged.go.emap) + ggtitle("GOBP: x vs Aged")
overlap.go.emapplot <- emapplot(overlap.go.emap) + ggtitle("GOBP: Overlap")

aged_vs_young.go.emapplot
x_vs_aged.go.emapplot
overlap.go.emapplot
```

```{r}
ggsave("./ORA/aged_vs_young_ora_go_emap.png", aged_vs_young.go.emapplot, width = 10, height = 10, bg = 'white')
ggsave("./ORA/x_vs_aged_ora_go_emap.png", x_vs_aged.go.emapplot, width = 10, height = 10, bg = 'white')
ggsave("./ORA/overlap_ora_go_emap.png", overlap.go.emapplot, width = 10, height = 10, bg = 'white')
```

### Cnet plot

```{r}
# Convert to readable gene symbols
aged_vs_young.go.symbol <- setReadable(aged_vs_young.go, OrgDb = org.Mm.eg.db, keyType = "ENTREZID")
x_vs_aged.go.symbol <- setReadable(x_vs_aged.go, OrgDb = org.Mm.eg.db, keyType = "ENTREZID")
overlap.go.symbol <- setReadable(overlap.go, OrgDb = org.Mm.eg.db, keyType = "ENTREZID")

# Visualize cnet
aged_vs_young.go.cnetplot <- cnetplot(aged_vs_young.go.symbol, categorySize="pvalue", color.params = list(foldChange=aged_vs_young.gene_list))
x_vs_aged.go.cnetplot <- cnetplot(x_vs_aged.go.symbol, categorySize="pvalue", color.params = list(foldChange=x_vs_aged.gene_list))
overlap.go.cnetplot <- cnetplot(overlap.go.symbol, categorySize="pvalue", color.params = list(foldChange=overlapping_genes))

aged_vs_young.go.cnetplot <- aged_vs_young.go.cnetplot + ggtitle("GO: Aged vs Young")
x_vs_aged.go.cnetplot <- x_vs_aged.go.cnetplot + ggtitle("GO: x vs Aged")
overlap.go.cnetplot <- overlap.go.cnetplot + ggtitle("GO: Overlap")

aged_vs_young.go.cnetplot
x_vs_aged.go.cnetplot
overlap.go.cnetplot
```

```{r}
ggsave("./ORA/aged_vs_young_ora_go_cnet.png", aged_vs_young.go.cnetplot, width = 10, height = 10, bg = 'white')
ggsave("./ORA/x_vs_aged_ora_go_cnet.png", x_vs_aged.go.cnetplot, width = 10, height = 10, bg = 'white')
ggsave("./ORA/overlap_ora_go_cnet.png", overlap.go.cnetplot, width = 10, height = 10, bg = 'white')
```

### Upset plot

```{r}
upsetplot(aged_vs_young.go)
upsetplot(x_vs_aged.go)
upsetplot(overlap.go)
```

## Reactome

```{r}
minGSSize = 15
maxGSSize = 300

# Obtain Reactome results
aged_vs_young.reactome <- enrichPathway(
  gene = names(aged_vs_young.gene_list),
  organism = 'mouse',
  pvalueCutoff = 0.05,
  minGSSize = minGSSize,
  maxGSSize = maxGSSize)

x_vs_aged.reactome <- enrichPathway(
  gene = names(x_vs_aged.gene_list),
  organism = 'mouse',
  pvalueCutoff = 0.05,
  minGSSize = minGSSize,
  maxGSSize = maxGSSize)

overlap.reactome <- enrichPathway(
  gene = overlapping_genes,
  organism = 'mouse',
  pvalueCutoff = 0.05,
  minGSSize = minGSSize,
  maxGSSize = maxGSSize)

# Visualize Reactome results
aged_vs_young.reactome.dotplot <- dotplot(aged_vs_young.reactome) + ggtitle("Reactome: Aged vs Young")
x_vs_aged.reactome.dotplot <- dotplot(x_vs_aged.reactome) + ggtitle("Reactome: x vs Aged")
overlap.reactome.dotplot <- dotplot(overlap.reactome) + ggtitle("Reactome: Overlap")

aged_vs_young.reactome.dotplot
x_vs_aged.reactome.dotplot
overlap.reactome.dotplot
```

```{r}
ggsave("./ORA/aged_vs_young_ora_reactome.png", aged_vs_young.reactome.dotplot, width = 8, height = 8, bg = 'white')
ggsave("./ORA/x_vs_aged_ora_reactome.png", x_vs_aged.reactome.dotplot, width = 8, height = 8, bg = 'white')
ggsave("./ORA/overlap_ora_reactome.png", overlap.reactome.dotplot, width = 8, height = 8, bg = 'white')
```

## Hallmark

```{r}
msigdb_data <- msigdbr(species = "Mus musculus", collection = "H")
msigdb_data.entrez <- dplyr::select(msigdb_data, gs_name, ncbi_gene)

minGSSize = 15
maxGSSize = 300

# Obtain Hallmark results
aged_vs_young.hallmark <- enricher(gene = names(aged_vs_young.gene_list), TERM2GENE = msigdb_data.entrez, minGSSize = minGSSize, maxGSSize = maxGSSize)
x_vs_aged.hallmark <- enricher(gene = names(x_vs_aged.gene_list), TERM2GENE = msigdb_data.entrez, minGSSize = minGSSize, maxGSSize = maxGSSize)
overlap.hallmark <- enricher(gene = overlapping_genes, TERM2GENE = msigdb_data.entrez, minGSSize = minGSSize, maxGSSize = maxGSSize)

# Visualize Hallmark results
aged_vs_young.hallmark.dotplot <- dotplot(aged_vs_young.hallmark) + ggtitle("Hallmark: Aged vs Young")
x_vs_aged.hallmark.dotplot <- dotplot(x_vs_aged.hallmark) + ggtitle("Hallmark: x vs Aged")
overlap.hallmark.dotplot <- dotplot(overlap.hallmark) + ggtitle("Hallmark: Overlap")

aged_vs_young.hallmark.dotplot
x_vs_aged.hallmark.dotplot
overlap.hallmark.dotplot
```

```{r}
ggsave("./ORA/aged_vs_young_ora_hallmark.png", aged_vs_young.hallmark.dotplot, width = 8, height = 8, bg = 'white')
ggsave("./ORA/x_vs_aged_ora_hallmark.png", x_vs_aged.hallmark.dotplot, width = 8, height = 8, bg = 'white')
ggsave("./ORA/overlap_ora_hallmark.png", overlap.hallmark.dotplot, width = 8, height = 8, bg = 'white')
```

# Gene Set Enrichment Analysis (GSEA)

## Load databases

```{r}
# Load gene sets (password and registration required -> download manually from https://www.gsea-msigdb.org/gsea/msigdb/mouse/collections.jsp)
GO <- read.gmt("./Supplement/Databases/m5.go.v2024.1.Mm.symbols.gmt")
GOBP <- read.gmt("./Supplement/Databases/m5.go.bp.v2024.1.Mm.symbols.gmt")
Reactome <- read.gmt("./Supplement/Databases/m2.all.v2024.1.Mm.symbols.gmt")
Hallmark <- read.gmt("./Supplement/Databases/mh.all.v2024.1.Mm.symbols.gmt")
```

## Create ranked gene list (Check Functions)

```{r}
# ranked_gene_list(df) df = DESeq2 results as df
aged_vs_young.ranked_genes <- ranked_gene_list(aged_vs_young.df)
x_vs_aged.ranked_genes <- ranked_gene_list(x_vs_aged.df)
```

## Perform GSEA

```{r}
GENE_SET = GOBP
minGSSize = 15
maxGSSize = 300

aged_vs_young.GSEA <- GSEA(
    aged_vs_young.ranked_genes,
    TERM2GENE = GENE_SET,
    minGSSize = minGSSize,
    maxGSSize = maxGSSize,
    verbose = FALSE,
    eps = 0
  )

x_vs_aged.GSEA <- GSEA(
    x_vs_aged.ranked_genes,
    TERM2GENE = GENE_SET,
    minGSSize = minGSSize,
    maxGSSize = maxGSSize,
    verbose = FALSE,
    eps = 0
  )
```

## Create interactive data table

```{r}
# Transform GSEA results into dataframe
aged_vs_young.GSEA.df <- as_tibble(aged_vs_young.GSEA@result)
x_vs_aged.GSEA.df <- as_tibble(x_vs_aged.GSEA@result)

aged_vs_young.GSEA.datatable <- datatable(aged_vs_young.GSEA.df,
                                                extensions = c('KeyTable', 'FixedHeader'),
                                                caption = 'Signatures enriched by Aging',
                                                options = list(keys = TRUE, searchHighlight = TRUE, pageLength = 10, lengthMenu = c('10', '25', '50', '100')))

aged_vs_young.GSEA.datatable

x_vs_aged.GSEA.datatable <- datatable(x_vs_aged.GSEA.df,
                                                extensions = c('KeyTable', 'FixedHeader'),
                                                caption = 'Signatures enriched by x',
                                                options = list(keys = TRUE, searchHighlight = TRUE, pageLength = 10, lengthMenu = c('10', '25', '50', '100')))

aged_vs_young.GSEA.datatable
```

## Obtain top 20 GSEA results (Check Functions)

```{r}
# ranked_gene_list(res.GSEA) res.GSEA  = GSEA results
aged_vs_young.GSEA.top_20 <- gsea_top20(aged_vs_young.GSEA)
x_vs_aged.GSEA.top_20 <- gsea_top20(x_vs_aged.GSEA)
```

## Plot top 20 GSEA results

```{r}
# Aged vs young
aged_vs_young.GSEA.barplot <- ggplot(
  aged_vs_young.GSEA.top_20,
  aes(
    x = NES,
    y = ID,
    fill = direction,
    alpha = neglog10_padj
  )
) +
  geom_col(width = 0.7) +
  scale_fill_manual(
    values = c(
      positive = "#FF4747",
      negative = "steelblue"
    )
  ) +
  scale_alpha_continuous(range = c(0.4, 1)) +
  theme_bw() +
  labs(
    title = paste0(
      "Top 10 Positive and Negative NES (Hallmark)\nin Aged vs Young"
    ),
    x = "Normalized Enrichment Score (NES)",
    y = "",
    fill = "Direction",
    alpha = expression(-log[10](adj.~p))
  )

aged_vs_young.GSEA.barplot

# x vs aged
x_vs_aged.GSEA.barplot <- ggplot(
  x_vs_aged.GSEA.top_20,
  aes(
    x = NES,
    y = ID,
    fill = direction,
    alpha = neglog10_padj
  )
) +
  geom_col(width = 0.7) +
  scale_fill_manual(
    values = c(
      positive = "#FF4747",
      negative = "steelblue"
    )
  ) +
  scale_alpha_continuous(range = c(0.4, 1)) +
  theme_bw() +
  labs(
    title = paste0(
      "Top 10 Positive and Negative NES (Hallmark)\nin x vs Aged"
    ),
    x = "Normalized Enrichment Score (NES)",
    y = "",
    fill = "Direction",
    alpha = expression(-log[10](adj.~p))
  )

x_vs_aged.GSEA.barplot
```

```{r}
dir.create("./GSEA")

ggsave("./GSEA/aged_vs_young_gsea_hallmark.png", aged_vs_young.GSEA.barplot, width = 10, height = 6, bg = 'white')
ggsave("./GSEA/x_vs_aged_gsea_hallmark.png", x_vs_aged.GSEA.barplot, width = 10, height = 6, bg = 'white')
```

## Export GSEA results

```{r}
# Create workbook
wb2 <- createWorkbook()

# List GSEA data frames
GSEA.df.list <- list(
  "Aged_vs_Young" = aged_vs_young.GSEA.df,
  "x_vs_Aged" = x_vs_aged.GSEA.df
)

# Loop through the list and add each data frame as a new sheet
for (sheet_name in names(GSEA.df.list)) {
  addWorksheet(wb2, sheet_name)  # Create a new sheet
  writeData(wb2, sheet = sheet_name, GSEA.df.list[[sheet_name]])  # Write data
}

# Save excel file
saveWorkbook(wb2, "./GSEA/GSEA_results.xlsx", overwrite = TRUE)
```

# DEG barplots (from selected gene sets)

```{r}
GENE_SET = GOBP
PATHWAY_NAME <- "GOBP_CORONARY_VASCULATURE_DEVELOPMENT"

genes_in_pathway <- GENE_SET$gene[GENE_SET$term == PATHWAY_NAME]

# Filter DESeq2 results for genes in this pathway
pathway_results <- x_vs_aged[rownames(x_vs_aged) %in% genes_in_pathway, ]

# Filter, sort, convert to data frame, and remove NA rows
pathway_results.df <- x_vs_aged %>%
  as.data.frame() %>%
  rownames_to_column(var = "gene") %>%
  filter(gene %in% genes_in_pathway) %>%
  arrange(desc(log2FoldChange)) %>%
  filter(padj < 0.05) %>%
  filter(abs(log2FoldChange) > 0.5) %>%
  na.omit()

# Convert gene column to a factor to preserve the order in the plot
pathway_results.df <- pathway_results.df %>%
  mutate(gene = factor(gene, levels = gene[order(log2FoldChange)]))

# Create the bar plot
pathway_results.barplot <- ggplot(pathway_results.df, aes(x = log2FoldChange, y = gene, fill = padj)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_gradient(low = "#FF4747", high = "azure", na.value = "azure") +  # Adjust gradient colors for padj
  labs(
    x = "log2 Fold Change",
    y = "",
    fill = "Adjusted p-value",
    title = paste0(PATHWAY_NAME, " (x vs Aged)")
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 10),  # Adjust y-axis gene label size
    axis.text.x = element_text(size = 10),  # Adjust x-axis text size
    legend.position = "right"
  )

pathway_results.barplot
```

```{r}
ggsave("./DEGs/x_vs_aged_barplot_gobp_coronary_vasculature_development.png", pathway_results.barplot, width = 8, height = 8, bg = 'white')
```
