---
title: "Ageing + X bulk mRNAseq (Transcription factor analysis)"
format: html
author: Tamas Lakat
---

# Experiment

The dataset contains bulk mRNA-sequencing data obtained from mouse heart samples (Young Control (YC; n=4), Aged Control (AC; n=4), Aged Treated (X; n=10))

# Preprocessing pipeline

1.  **Trimming + QC**
    - Software: fastp
    - Folders: 0_raw ->  1_trimmed
2.  **QC**
    - Software: fastqc, multiqc
    - Folders: 1_trimmed -> 2_fastqc
3.  **Pseudoalignment**
    - Software: Kallisto
    - Reference transcriptome version: Mus_musculus.GRCm39
    - Folders: 2_fastqc + 3_reference -> 4_aligned

# Install libraries

```{r}
#| label: Install libraries
#| include: false


if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install("decoupleR")
BiocManager::install("OmnipathR")
```

# Import libraries

```{r}
#| label: Load libraries
#| include: false

library(decoupleR)
library(OmnipathR)
library(tidyr)
library(tibble)
library(dplyr)
library(ggplot2)
library(ComplexHeatmap)
```

# DecoupleR

## Load data

```{r}
dds <- readRDS("./EDA/dds.Rds")
```

## TF activity heatmap

```{r}
vsd <- vst(dds, blind = TRUE)
mat <- assay(vsd)

net <- decoupleR::get_collectri(organism = 'mouse', 
                                split_complexes = FALSE)

sample_acts <- decoupleR::run_ulm(mat = mat, 
                                  net = net, 
                                  .source = 'source', 
                                  .target = 'target',
                                  .mor = 'mor', 
                                  minsize = 5)

sample_acts

n_tfs <- 25

# Transform to wide matrix
sample_acts_mat <- sample_acts %>%
  tidyr::pivot_wider(id_cols = 'condition', 
                     names_from = 'source',
                     values_from = 'score') %>%
  tibble::column_to_rownames('condition') %>%
  as.matrix()

# Get top tfs with more variable means across clusters
tfs <- sample_acts %>%
  dplyr::group_by(source) %>%
  dplyr::summarise(std = stats::sd(score)) %>%
  dplyr::arrange(-abs(std)) %>%
  head(n_tfs) %>%
  dplyr::pull(source)

sample_acts_mat <- sample_acts_mat[,tfs]

# Scale per sample
sample_acts_mat <- scale(sample_acts_mat)

# Choose color palette
colors <- rev(RColorBrewer::brewer.pal(n = 11, name = "RdBu"))
colors.use <- grDevices::colorRampPalette(colors = colors)(100)

my_breaks <- c(seq(-2, 0, length.out = ceiling(100 / 2) + 1),
               seq(0.05, 2, length.out = floor(100 / 2)))

dir.create("./TF")

# Plot

pheatmap::pheatmap(mat = sample_acts_mat,
                   color = colors.use,
                   border_color = "white",
                   breaks = my_breaks,
                   cellwidth = 15,
                   cellheight = 15,
                   treeheight_row = 20,
                   treeheight_col = 20)

# Save heatmap
pheatmap::pheatmap(mat = sample_acts_mat,
                   color = colors.use,
                   border_color = "white",
                   breaks = my_breaks,
                   treeheight_row = 20,
                   treeheight_col = 20,
                   filename = "TF/transcription_factors_heatmap.png",
                   width = 10,
                   height = 8 
)
```

## TF activity barplot

```{r}
# Load pairwise comparison data
aged_vs_young <- results(dds, contrast=c("group", "Aged Control", "Young Control"))
x_vs_aged <- results(dds, contrast=c("group", "Aged Treated", "Aged Control"))

# Convert to df
aged_vs_young.df <- as.data.frame(aged_vs_young)
aged_vs_young.df <- aged_vs_young.df[!is.na(aged_vs_young.df$padj), ]

x_vs_aged.df <- as.data.frame(x_vs_aged)
x_vs_aged.df <- x_vs_aged.df[!is.na(x_vs_aged.df$padj), ]

# Run ulm
DF = x_vs_aged.df
contrast_acts <- decoupleR::run_ulm(mat = DF[, 'stat', drop = FALSE], 
                                    net = net, 
                                    .source = 'source', 
                                    .target = 'target',
                                    .mor='mor', 
                                    minsize = 5)

# Filter top TFs in both signs
f_contrast_acts <- contrast_acts %>%
  dplyr::mutate(rnk = NA)

msk <- f_contrast_acts$score > 0

f_contrast_acts[msk, 'rnk'] <- rank(-f_contrast_acts[msk, 'score'])
f_contrast_acts[!msk, 'rnk'] <- rank(-abs(f_contrast_acts[!msk, 'score']))

tfs <- f_contrast_acts %>%
  dplyr::arrange(rnk) %>%
  head(n_tfs) %>%
  dplyr::pull(source)

f_contrast_acts <- f_contrast_acts %>%
  filter(source %in% tfs)

colors <- rev(RColorBrewer::brewer.pal(n = 11, name = "RdBu")[c(2, 10)])

tf.barplot <- ggplot2::ggplot(data = f_contrast_acts, 
                     mapping = ggplot2::aes(x = stats::reorder(source, score), 
                                            y = score)) + 
  ggplot2::geom_bar(mapping = ggplot2::aes(fill = score),
                    color = "black",
                    stat = "identity") +
  ggplot2::scale_fill_gradient2(low = colors[1], 
                                mid = "whitesmoke", 
                                high = colors[2], 
                                midpoint = 0) + 
  ggplot2::theme_minimal() +
  ggplot2::theme(axis.title = element_text(face = "bold", size = 12),
                 axis.text.x = ggplot2::element_text(angle = 45, 
                                                     hjust = 1, 
                                                     size = 10, 
                                                     face = "bold"),
                 axis.text.y = ggplot2::element_text(size = 10, 
                                                     face = "bold"),
                 panel.grid.major = element_blank(), 
                 panel.grid.minor = element_blank()) +
  ggplot2::xlab("X vs Aged TFs")

tf.barplot
```

```{r}
ggsave("./TF/aged_vs_young_tf_barplot.png", tf.barplot, width = 8, height = 4, bg = 'white')
```

```{r}
# Visualize genes belong to TF
tf <- 'Nr3c1'

df <- net %>%
  dplyr::filter(source == tf) %>%
  dplyr::arrange(target) %>%
  dplyr::mutate(ID = target, color = "3") %>%
  tibble::column_to_rownames('target')

inter <- sort(dplyr::intersect(rownames(DF), rownames(DF)))

df <- df[inter, ]

df[, c('logfc', 't_value', 'p_value')] <- DF[inter, c('log2FoldChange', 'stat', 'pvalue')]

df <- df %>%
  dplyr::mutate(color = dplyr::if_else(mor > 0 & t_value > 0, '1', color)) %>%
  dplyr::mutate(color = dplyr::if_else(mor > 0 & t_value < 0, '2', color)) %>%
  dplyr::mutate(color = dplyr::if_else(mor < 0 & t_value > 0, '2', color)) %>%
  dplyr::mutate(color = dplyr::if_else(mor < 0 & t_value < 0, '1', color))

df_naomit <- na.omit(df)

colors <- rev(RColorBrewer::brewer.pal(n = 11, name = "RdBu")[c(2, 10)])

p2 <- ggplot2::ggplot(data = df, 
                     mapping = ggplot2::aes(x = logfc, 
                                            y = -log10(p_value), 
                                            color = color,
                                            size = abs(mor))) + 
  ggplot2::geom_point(size = 2.5, 
                      color = "black") + 
  ggplot2::geom_point(size = 1.5) +
  ggplot2::scale_colour_manual(values = c(colors[2], colors[1], "grey")) +
  ggrepel::geom_label_repel(mapping = ggplot2::aes(label = ID,
                                                   size = 1)) + 
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "none") +
  ggplot2::geom_vline(xintercept = 0, linetype = 'dotted') +
  ggplot2::geom_hline(yintercept = 0, linetype = 'dotted') +
  ggplot2::ggtitle(tf)

p2
```

```{r}
ggsave("./TF/volcano_x_vs_aged_Nr3c1.png", p2, width = 8, height = 8, bg = 'white')
```

## Target genes heatmap

```{r}
# Get target name vector
nr3c1_targets <- net$target[net$source=='Nr3c1']
valid_targets <- intersect(nr3c1_targets, rownames(vsd))

# Get heatmap matrix
nr3c1_mat <- assay(vsd[valid_targets, ])
nr3c1_mat <- t(scale(t(nr3c1_mat)))

# Ensure only available samples are used
desired_order <- intersect(desired_order, colnames(nr3c1_mat))

# Reorder the columns in the matrix based on the specified order
nr3c1_mat <- nr3c1_mat[, desired_order]

# Get color vectors
nr3c1_sig.df <- scatter_data.merged[scatter_data.merged$gene %in% rownames(nr3c1_mat), ]
rownames(nr3c1_sig.df) <- nr3c1_sig.df$gene

row_sig_labels <- scatter_data.merged$significance[match(rownames(nr3c1_mat), scatter_data.merged$gene)]
row_sig_labels[is.na(row_sig_labels)] <- "0"

row_color_map <- c("0" = "grey", "Aged" = "lightblue", "x" = "steelblue", "Both" = "#FF4747")
row_colors <- row_color_map[row_sig_labels]
row_colors[is.na(row_colors)] <- "grey"

# Create heatmap with your preferred styling
nr3c1_mat.heatmap <- Heatmap(
  nr3c1_mat, 
  cluster_rows = TRUE, 
  cluster_columns = FALSE, 
  col = colorRamp2(c(-2, 0, 2), c("steelblue", "white", "#FF4747")),
  show_row_dend = FALSE,
  show_column_dend = FALSE,
  show_heatmap_legend = TRUE,
  heatmap_legend_param = list(
    title = "z-score",  
    title_gp = gpar(fontsize = 14),  
    labels_gp = gpar(fontsize = 12)  
  ),
  rect_gp = gpar(col = "white", lwd = 1),
  column_names_gp = gpar(fontsize = 10),
  row_names_gp = gpar(fontsize = 10, col = row_colors), 
  row_names_max_width = unit(35, "cm")
)

nr3c1_mat.heatmap
```

```{r}
tryCatch({
    # Open the device
    png("TF/Nr3c1_heatmap.png", 
        width = 1600,
        height = 4000,
        res = 300)
    
    # Calculate and Draw
    print("Drawing heatmap...")
    ComplexHeatmap::draw(nr3c1_mat.heatmap)
    
    # Force the device to save
    dev.off()
    print("File saved successfully!")
    
}, error = function(e) {
    print(paste("Error caught:", e$message))
    if(!is.null(dev.list())) dev.off()
})
```

### 'Both' target genes heatmap

```{r}
# Get Both genes
is_both <- row_sig_labels == "Both"
nr3c1_both_mat <- nr3c1_mat[is_both, ]

# Create heatmap
nr3c1_both.heatmap <- Heatmap(
  nr3c1_both_mat, 
  cluster_rows = TRUE, 
  cluster_columns = FALSE, 
  col = colorRamp2(c(-2, 0, 2), c("steelblue", "white", "#FF4747")),
  show_row_dend = FALSE,
  show_column_dend = FALSE,
  show_heatmap_legend = TRUE,
  heatmap_legend_param = list(
    title = "z-score",  
    title_gp = gpar(fontsize = 14),  
    labels_gp = gpar(fontsize = 12)  
  ),
  rect_gp = gpar(col = "white", lwd = 1),
  column_names_gp = gpar(fontsize = 10),
  row_names_gp = gpar(fontsize = 10), 
  row_names_max_width = unit(35, "cm"),
  column_title = "Nr3c1 Targets: Significant in Both"
)

nr3c1_both.heatmap
```

```{r}
tryCatch({
    # Open the device
    png("TF/Nr3c1_both_heatmap.png", 
        width = 2000,
        height = 1800,
        res = 300)
    
    # Calculate and Draw
    print("Drawing heatmap...")
    ComplexHeatmap::draw(nr3c1_both.heatmap)
    
    # Force the device to save
    dev.off()
    print("File saved successfully!")
    
}, error = function(e) {
    print(paste("Error caught:", e$message))
    if(!is.null(dev.list())) dev.off()
})
```
